Follow-up Samples

### 2017-07-22 submission/manuscript.Rmd

Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota. If the microbial community has an affect on recurrence risk or tumorigenesis it would be reasonable to expect treatment to remove a lesion to affect the microbiota. To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner? If so, does the treatment alter the community to more closely resemble that of individuals with normal colons?

@@

Our study focused on comparing the microbiota of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment. For all three groups of patients, we observed changes in their microbiota. Some of these changes, specifically for adenoma and advanced adenoma, may be due to normal temporal variation but those with CRC clearly had large microbiota changes that would be greater than what could be expected from this type of variation. After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups. This change resulted in communities that more closely resembled those of patients with a normal colon.  This may suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at reducing the associated bacterial communities.  Understanding the effect of treatment on the microbiota of those diagnosed with carcinomas may have important implications for reducing disease recurrence.  It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to assess the risk of recurrence.

@@

This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions. Although we cannot make a specific statment regarding the effect of a specific treatment on the gut microbiota, due to the low number of samples, there are still a number of exciting new avenues to explore. One of the most exciting of these future directions is the possibility that markers within the microbiota could be used to potentially evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma. If such an approach is effective, it might be possible to target the microbiota as part of adjuvant therapy, if the biomarkers identified play a key role in the disease process. Our data provides additional evidence on the importance of the microbiota in tumorigenesis by addressing the recovery of the microbiota after treatment and opens interesting avenues of research into how these changes may affect recurrence.

@@

***Study Design and Patient Sampling.***
Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016]. Briefly, samples were stored on ice for at least 24h before freezing. Although we cannot exclude that this sampling protocol may have impacted the gut microbiota composition all samples were subjected to the same methodology. Study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis. Samples used to build the models for prediction were collected either prior to a colonoscopy or between one and two weeks after initial colonoscopy. The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013]. Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 1]. Patients were diagnosed by colonoscopic examination and histopathological review of any biopsies taken. Patients were classified as having advanced adenoma if they had an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology. This study was approved by the University of Michigan Institutional Review Board. All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***Treatment***
The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surigically [Table 1]. The type of chemotherapy used for patients with CRC varied markably. The types used included Oxaliplatin, Levicovorin, Folfox, Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin. These were used individually or in combination with others depending on the patient [Table 1]. If an individual was treated with radiation they were also always treated with chemotherapy. Radiation therapy generally used 18 mV photons for treatment.

@@

Similarily, the Random Forest [@breiman_random_2001] algorithm was also used to create the three models used to classify normal versus diagnosis. These samples were obtained using the same methodology as described earlier in this section. All samples used for this component of model training were from pre-treatment samples. The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2]. There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` individuals in the normal versus carcinoma model [Table 2].

@@

**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs. B) Advanced Adenoma OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10 MDA value obtained from 100 different 80/20 splits while the lighter colored circles represents the value obtained for a specific run.

@@

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time, Surgery, chemo_received, radiation_received)

t2 <- as.data.frame(matrix(nrow = 9, ncol = 3, 
                           dimnames = 
                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time",  
                                           "Surgery_only", "Surg_Chemo", "Surg_Chemo_Radiation"), 
                                  coln = c("Adenoma", "SRN", "Carcinoma"))))

@@

  t2["Dx_Bin", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
  t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "female") %>% select(n)
  t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
  t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100 
  t2["post_time", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(time, na.rm = TRUE))
  t2["Surgery_only", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & Surgery == "Y" & chemo_received == "no") %>% select(Surgery) %>% count()
  t2["Surg_Chemo", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & chemo_received == "yes" & radiation_received == "no") %>% select(chemo_received) %>% count()
  t2["Surg_Chemo_Radiation", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & radiation_received == "yes") %>% select(radiation_received) %>% count()
}

@@

  age_sd <- c(age_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
  BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
  post_time_sd <- c(post_time_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(time, na.rm = TRUE)))[, "std"])
}

@@

ttemptable <- as.data.frame(t(t2)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(age_sd, digits = 2, nmsall = 2)), 
                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2), 
                                        BMI = paste(format(BMI, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(BMI_sd, digits = 2, nmsall = 2)), 
                                        White = format(White, digits =2, nsmall = 2), 
                                        post_time = paste(format(post_time, digits = 2, nsmall = 2),
                                                    label = "\u00B1",
                                                    format(post_time_sd, digits = 2, nmsall = 2)), 
                                        Surgery_only = format(Surgery_only, digits = 2, nsmall = 0), 
                                        Surg_Chemo = format(Surg_Chemo, digits = 2, nsmall = 0), 
                                        Surg_Chemo_Radiation = format(Surg_Chemo_Radiation, digits = 2, nsmall = 0))

ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
             ttemptable$BMI, ttemptable$White, ttemptable$post_time, ttemptable$Surgery_only, ttemptable$Surg_Chemo, ttemptable$Surg_Chemo_Radiation)

@@

colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label, 
                   "Surgery Only", "Surgery & Chemotherapy", "Surgery, Chemotherapy, & Radiation")

>>>>

Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota. If the microbial community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would affect the microbiota and risk of recurrence. To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner? If so, does the treatment alter the community to more closely resemble that of individuals with normal colons?

@@

Our study focused on comparing the microbiota of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment. For all three groups of patients, we observed changes in their microbiota. After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups. This change resulted in communities that more closely resembled those of patients with a normal colon.  This may suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at reducing the associated bacterial communities.  Understanding the effect of treatment on the microbiota of those diagnosed with carcinomas may have important implications for reducing disease recurrence.  It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to assess the risk of recurrence.

@@

This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions. The most exciting future direction from the current study is the possibility that markers within the microbiota could be used to evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma. If such an approach is effective, it might be possible to target the microbiota as part of adjuvant therapy. Our data provides additional evidence on the importance of the microbiota in tumorigenesis by addressing the recovery of the microbiota after treatment and opens interesting avenues of research into how these changes may affect recurrence.

@@

***Study Design and Patient Sampling.***
Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016]. Briefly, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis. Samples used to build the models for prediction were collected either prior to a colonoscopy or between one and two weeks after initial colonoscopy. The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013]. Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 1]. Patients were diagnosed by colonoscopic examination and histopathological review of any biopsies taken. Patients were classified as having advanced adenoma if they had an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology. This study was approved by the University of Michigan Institutional Review Board. All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

@@

Similarily, the Random Forest [@breiman_random_2001] algorithm was also used to create the three models used to classify normal versus diagnosis. The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2]. There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` individuals in the normal versus carcinoma model [Table 2].

@@

**Figure 2: The 10 OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs. B) Advanced Adenoma OTUs. C) Carcinoma OTUs.

@@

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time)

t2 <- as.data.frame(matrix(nrow = 6, ncol = 3,
                           dimnames =
                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time"),
                                  coln = c("Adenoma", "SRN", "Carcinoma"))))

@@

  t2["Dx_Bin", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
  t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "female") %>% select(n)
  t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
  t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100

  t2["post_time", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(time, na.rm = TRUE))

}

@@

  age_sd <- c(age_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
  BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
  post_time_sd <- c(post_time_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(time, na.rm = TRUE)))[, "std"])

}

@@

ttemptable <- as.data.frame(t(t2)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2),
                                                    label = "\u00B1",
                                                    format(age_sd, digits = 2, nmsall = 2)),
                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2),
                                        BMI = paste(format(BMI, digits = 2, nsmall = 2),
                                                    label = "\u00B1",
                                                    format(BMI_sd, digits = 2, nmsall = 2)),
                                        White = format(White, digits =2, nsmall = 2),
                                        post_time = paste(format(post_time, digits = 2, nsmall = 2),
                                                    label = "\u00B1",
                                                    format(post_time_sd, digits = 2, nmsall = 2)))

ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f,
             ttemptable$BMI, ttemptable$White, ttemptable$post_time)

@@

colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label)
